---
title: "Algorithmic Problems"
---

## LeetCode in R?...

This page consists of algorithmic problems implemented in R, for fun and an 
extra challenge. Nothing here is in scope for STAT 33B. If you think you've 
solved some fun problem successfully in R, please send it to me! I'd be happy 
to include it here for others to see. 

If you're looking for problems to solve, I'd recommend [Codeforces](https://codeforces.com/problemset) 
or [LeetCode](https://leetcode.com/problemset/all/). 
Unfortunately, neither has R support. For this, I'd recommend [CodeChef](https://www.codechef.com/).

## Two Sum

To start, the most popular problem on LeetCode! Click [here](https://leetcode.com/problems/two-sum/) to read the problem.

```{r}
two_sum <- function(nums, target) {
  # Your code here
}
```

```{r}
#| code-fold: true
#| code-summary: Show Solution
two_sum <- function(nums, target) {
  num_indices <- new.env() # we use an environment as a hash map
  for (i in seq_along(nums)) {
    dif <- as.character(target - nums[i])
    dif_index <- num_indices[[dif]]
    if (!is.null(dif_index))
      return(c(i, dif_index))
    num_indices[[as.character(nums[i])]] <- i
  }
  -1
}
```

_This page is in progress. Contributions are welcome._