---
title: "Lab 01: Intro, Setup"
format: 
  revealjs:
    footer: "[Return to Home Page](https://red-robby.github.io/)"
    chalkboard: true
---

## Welcome!

Welcome to STAT 133! I'll be your uGSI, **Jakob Sorensen**.

::: incremental
-   Contact: jakobsorensen\@berkeley.edu

-   My Sections:

    -   11 AM to 1 PM  in *Evans 340*

    -   3 PM to 5  PM in *Evans 340*

-   Office Hours: TBD
:::  

## Expectations

- Weekly Lab (13 total)
- Timed Quizzes (2 total -- via bCourses)
- Biweekly[^1] -- every 2 weeks -- Homework (6 total)
- Final Exam

[^1]: Isn't it unfortunate _biweekly_ can mean twice a week _and_ every 2 weeks?

## Our Lab

- For the **first 15-40 minutes**, I'll lecture briefly on select topics from the week's 
material and/or we'll solve a few challenges together. 

- Afterwards, you are free to work on the lab, and I'm available to help. 


## My Notes

I'll be hosting the materials I create for the course at the following website:

::: {.r-fit-text}
[https://red-robby.github.io/]()[^2]
:::

. . .

_You can also find my materials from STAT 33B, if you're interested._

[^2]: Sorry for the ugly name! It'll be linked on bCourses. 

## Installing R and RStudio

Please install R and RStudio now if you haven't already.

::: {.r-fit-text}
-   **R**: [https://www.r-project.org/]()

-   **RStudio:** [https://www.rstudio.com/products/rstudio/]()
:::

## RStudio Demo

Separate window. 

Demo in separate window. 

## RMarkdown

RMarkdown (`.Rmd`) is file type that allows us to merge,

-   Formatted written material,
-   and **R** code

into a single document.

::: {.fragment .fade-right fragment-index="1"}
::: {.fragment .highlight-blue fragment-index="1"}
*This presentation is using a variant of RMarkdown!*
:::
:::

## *This* is RMarkdown?

::: {.fragment .grow}
```{rmd, echo=TRUE}
## RMarkdown

RMarkdown (``.Rmd``) is file type that allows us to merge,

-   Formatted written material,
-   and **R** code

into a single document.
```
:::

. . .

## RMarkdown - Basic Formatting

Software like Microsoft Word and Google Docs **hides** the underlying format representation.

::: {.fragment .fade-right fragment-index="1"}
::: {.fragment .highlight-blue fragment-index="1"}
In a markup language, we *embed the formatting alongside the text.*
:::
:::

::: {.fragment .fade-right}
**An Example: HTML (HyperText Markup Language)**

```{html, echo=TRUE}
This is <b>bold!</b> But this is merely <i>emphasized!</i>
```
:::

::: {.fragment .fade-left}
**Another Example: LaTeX**

```{LaTeX, echo=TRUE}
This is \textbf{bold!} But this is merely \textit{emphasized!}
```
:::

::: {.fragment .fade-up}
**RMarkdown**

```{rmd, echo=TRUE}
This is **bold!** But this is merely _emphasized!_
```
:::

## Headers

### Aren't headers...

#### ... just lovely?

::: {.fragment .fade-right}
```{rmd, echo=TRUE}
## Headers

### Aren't headers...

#### ... just lovely?
```
:::

::: {.fragment .fade-up}
$N$ `#` characters denotes an $Nth$ level header.
:::

## Font Face: **Bold**

I'm **bold** and **uncontrolled!**

::: {.fragment .fade-right}
```{rmd, echo=TRUE}
## Font Face: **Bold**

I'm **bold** and **uncontrolled!**
```
:::

::: {.fragment .fade-up}
Surrounding text with `**` at each end makes it **bold**.
:::

## Font Face: *Italics*

Don't eat *that* burger! Don't *eat* that burger!

::: {.fragment .fade-right}
```{rmd, echo=TRUE}
## Font Face: _Italics_

Don't eat _that_ burger! Don't _eat_ that burger!
```
:::

::: {.fragment .fade-up}
Surrounding text with `_` at each end *italicizes* it.
:::

## Font Face: `Plain Text or Code`

The variable `x` passed into function `main`.

::: {.fragment .fade-right}
```{rmd, echo=TRUE}
## Font Face: ``Plain Text or Code``

The variable ``x`` passed into function ``main``.
```
:::

::: {.fragment .fade-up}
Surrounding text with ``` `` ``` at each end makes it plain text (like `code`).
:::

## And more...

We can also embed lists, tables, images, and many other components!

## Code Chunks

We include code in what we call *code chunks*. A basic example in this document:

::: {.fragment .fade-right fragment-index="1"}
```{r}
#| echo: fenced
print("Hello, World!")

x <- 3
y <- 5
print(paste("x + y =", x + y))
```
:::

::: {.fragment .fade-right fragment-index="2"}
::: {.fragment .highlight-blue fragment-index="2"}
*Note that we've included the output of our code directly in the document.*
:::
:::

## Code Chunk: Plot Example

::: {.fragment .fade-up}
```{r}
#| echo: fenced
plot(1:10, seq(10, 100, 10), type = "l", 
     xlab = "Time Spent Studying (Hours)", ylab = "Quiz Grade (%)")
```
:::

## Knitting

Going from the plain text format (`.rmd`) to a presentation format (`.pdf`, `.html`) is called *knitting* the document.

## RMarkdown Demo

Separate window.



